Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-14 manuscript/manuscript.tex
edits to decay results section

Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
All phylogenetic substitution models \skhcomment{all?} are stochastic matrices and, as such, reach stationary state. 
Stationary state describes the expected sequence composition after a very long evolutionary time and this stationary state sequence composition is independent to time. 
In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.

By definition, the expected sequence composition varies from site to site for the site-specific stationary state models (\ref{fig:decay} ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay} GY94, blue).
The biggest difference in stationary state between the models is at constrained sites, such as site 305 in \ref{fig:decay}. 
At such a site, the ExpCM would estimate a much longer branch than the GY94 given the observed divergence of a set of sequences.

Rate variation, as modeled by $\Gamma\omega$, affect the rate at which stationary state is approached by not the sequence composition at stationary state itself. 
Stationary state is a fundamental feature of stochastic matrices and is invariant to multiplicative terms, such as $\omega$. 
In \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical.

This means even complex modeling of the $\omega$ to account for purifying selection will not affect the expectation at evolutionary equilibrium. 
We inferred a site-specific $\omega$ value from the ExpCM \citep{spielman2015relationship} and applied these values to the GY94 model, GY94+$\Gamma\omega$ (\ref{fig:decay} yellow). 
At constrained sites, this model takes even longer to reach stationary state than the GY94+$\Gamma\omega$ but the eventual stationary state is identical to both the GY94 and GY94+$\Gamma\omega$ stationary state.

No matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.

>>>>

Given a single branch, a substitution model transforms sequence divergence into branch length, which is proportional to time under a molecular clock assumption. 
However, a complicating factor in this estimation procedure is the stationary state of the model. 
Stationary state, as the principle eigenvector of a stochastic matrix, is a fundamental feature of any substitution model. 
It describes the expected sequence composition after a very long evolutionary time. 
Additionally, the stationary state does not change under application of the transition matrix, which means this sequence composition remains constant as time increases. 
The invariance of the stationary state is represented by the long ``tails" in \ref{fig:decay}. 
The exact sequence composition of stationary state is unique given the parametrization of the model.

We visualized the effect of stationary state on branch length estimation by examining the expected amino-acid identity of a set of sequences given some branch length for site-specific and uniform stationary state models. 
By definition, the expected identity given a very long time varies from site to site for the site-specific stationary state models (\ref{fig:decay}, ExpCM, red) but is constant across sites for the uniform stationary state models (\ref{fig:decay}, GY94, blue). 
Due to the assumed equivalence of amino-acid preferences in the GY94 model, the biggest difference in stationary state between the ExpCM and the GY94 is seen at constrained sites, such as site 305 in \ref{fig:decay}. 
The ExpCM would estimate a much longer branch than the GY94 model given some observed divergence at such a site.

Modeling site-to-site rate variation by $\Gamma\omega$ affects the rate at which stationary state is approached but not the sequence composition at stationary state itself. 
For example, in \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical. 
Because the stationary state is invariant to multiplicative terms, increasingly complex modeling of $\omega$ will never fully recapitulate the difference in expected sequence composition between the ExpCM and the GY94 at constrained sites. 
We inferred an $\omega$ value from the ExpCM for each site \citep{spielman2015relationship} and applied this values to the GY94 model (\ref{fig:decay} yellow)
These site-specific values drastically decreased the amount of time it takes to reach stationary state at the constrained sites but still does not affect the stationary state. 
This means that no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.
